Features of Accretion-Phase Gravitational Wave Emission from Two-dimensional Rotating Core-Collapse Supernovae

### 2019-06-26 ms.tex
More text updates for ARXiV

\author{Michael A. Pajkos}
\affiliation{Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Computational Mathematics, Science, and Engineering, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Joint Institute for Nuclear Astrophysics-Center for the Evolution of the Elements, Michigan State University, East Lansing, MI 48824, USA}

@@

\author{Kuo-Chuan Pan}
\affiliation{Department of Physics and Institute of Astronomy, National Tsing Hua University, Hsinchu 30013, Taiwan}

@@

\section{Methods and Simulation Setup}
\label{sec:method}
We utilize the \texttt{FLASH} (version 4) multiscale, multiphysics adaptive mesh refinement simulation framework for our simulations \citep{fryxell:2000,dubey:2009}.\footnote{http://flash.uchicago.edu/site/}  We employ a modified, GR, effective potential \citep{marek:2006, oconnor:2018} incorporated into the multipole Poisson solver of \citet{couch:2013a}, where we retain spherical harmonic orders up through 16.   We utilize the SFHo EOS in all of our 15 simulations \citep{steiner:2013}.  Our grid setup is a 2D cylindrical geometry with the PARAMESH (v.4-dev) library for adaptive mesh refinement  \citep{macneice:2000}.  The outer boundary is $10^4$ km in all directions, with nine levels of refinement, yielding a finest grid spacing of about 0.65 km.
The maximum allowed level of refinement is decreased as a function of spherical radius, $r$, in order to maintain a resolution aspect ratio, $\Delta x_i / r$, of about 0.01, corresponding approximately to an ``angular'' resolution of $0.5^{\circ}$ .

@@

\begin{figure*}[t!]
  \centering     %%% not \center
  \includegraphics[width=\textwidth]{tbe6tbe300_combined_M1_long_referee.pdf}
  \caption{Amplitude spectral density (ASD) plot of all progenitors for all rotation rates from $t_{be}+6$ ms $\rightarrow t_{be}+300$ ms, \textcolor{red}{[10]with an assumed distance of 10 kpc.}  The rotational muting of the fundamental PNS g-mode is displayed as the peak frequency ($\sim 800$ Hz) becomes less prevalent, with increasing rotation rate.  Likewise, the low frequency signals ($\sim40$ Hz) from the gain region become more audible, with increasing rotational velocity.  The damping of the vibrational modes of the PNS allows the slower postshock convection to contribute more to the overall GW signal.  Plotted in the black dashed line is the design sensitivity curve for aLIGO in the zero detuning, high sensitivity configuration \citep{barsotti:2018}.  The cyan dashed line is the predicted KAGRA detuned, sensitivity curve \citep{komari:2017}.  The purple dashed line is the design sensitivity curve for AdV \citep{abbott:2018}.}
  \label{fig:spetra_long}
\end{figure*}

@@

Overlaid on our ASD plots is the expected sensitivity of future GW observatories.  
\textcolor{red}{[10]In the black, cyan, and purple dashed lines we have plotted the sensitivity curves of design sensitivity for aLIGO in the zero detuning, high sensitivity configuration, the predicted KAGRA detuned sensitivity curve, and design sensitivity for AdV, respectively \citep{komari:2017,barsotti:2018,abbott:2018}.}
These curves represent the incoherent sum of the principal noise sources to the best understanding of the respective collaborations.  While these curves do not guarantee the performance of the detectors, they act as good guides for their anticipated sensitivities, nonetheless.

Beyond the decreased prevalence of the peak frequency, an interesting trend emerges in Figure \ref{fig:spetra_long} as rotation increases.  
We separate the GW signals by region within the star.  The top row of Figure \ref{fig:spetra_long} corresponds to GWs originating from the inner 50 km of the supernova, and the GW signal in the bottom row originates from radial distances between 50 km and 150 km from the supernova center.  \textcolor{red}{[10]  In the top row, we note the first peak of emission, around 80 Hz, is independent of rotation.  We point to the bright, higher $v_\mathrm{aniso}$ region in Figure \ref{fig:vaniso} within the first 25 ms postbounce that is present for all rotational velocities.}
Focusing on the bottom row, we highlight a noticeable difference in the amplitude of the low frequency contributions, particularly around 40 Hz.  The nonrotating progenitors have undetectable low frequency signals for all three detectors, whereas rotating progenitors create measurable signals at low frequencies. 
This enhanced low-frequency signal may provide an observable feature that can help determine progenitor angular momentum information.

The amplitude of low frequency GWs in the 50 km to 150 km region of the supernova increases with rotational velocity, but this trend does not occur within the inner 50 km.  As such, we restrict the low frequency GW contribution to the gain region.  We note the two main physical mechanisms in this region correspond to postshock convection and the SASI.  While both mechanisms are reduced in strength due to rotational effects, they do not completely cease.  This fact is displayed in Figure \ref{fig:vaniso}, as the region between 50 km and 150 km is nonzero.  For the nonrotating case, the high convective velocities (bright yellow) create higher frequency GWs within the 100 km region of interest.  As rotation velocity increases, convective velocities decrease enough to cease exciting the vibrational modes of the PNS.  These slower convective flows thereby reduce the total amount of power produced by the GWs and push the peak GW frequency---from the gain region---to lower frequencies.  Performing an order of magnitude estimate on the source of the low frequency signal, from Figure \ref{fig:vaniso}, we find  $v_{\mathrm{aniso}} \sim 1 \times 10^9$ cm s$^{-1}$ for $\Omega_0 = 0$ rad s$^{-1}$ and $v_{\mathrm{aniso}} \sim 5 \times 10^8$ cm s$^{-1}$ for $\Omega_0 = 3$ rad s$^{-1}$.  As the region of interest is $\sim 10^7$ cm we yield an estimated frequency of emission $f_{\mathrm{low}}$ around $\sim 100$ Hz and $\sim 50$ Hz, respectively.  These quantitative frequency estimates are reflected in the ASD as the contribution from peak frequency ($\sim 100$ Hz) from the gain region decreases, while the contribution $\sim 40$ Hz increases.

@@

\textcolor{red}{[1] The strength of this project is its ability analyze GWs hundreds of milliseconds postbounce from multiple progenitors while accurately accounting for rotation and neutrinos.  The wide breadth of parameter space we examine allows us to reveal certain rotational effects on the GW signal in the context of a controlled study.}  We have explored the influence of rotation on the GW emission from CCSNe for four different progenitors and four different core rotational speeds.  
We point out that there exists a roughly linear relation between compactness, $\xi$, and differential rotation parameter, $A$, as defined in Equation (\ref{eq:omega}). 
Using this relation, we calculate appropriate $A$ values for each progenitor mass, based on their individual compactness parameters of the \citet{Suk:2016} progenitors.  Of our 15 simulations, only two nonrotating progenitors have average shock radii that show substantial shock expansion, while the remaining rotating progenitors do not because of rotationally inhibited convection in the gain region and less neutrino production.  In agreement with other recent work \citep[e.g.,][]{summa:2018}, we find a complex interplay between centrifugal support and neutrino heating as successful explosions do not display a monotonic relationship with rotation.

While there are more accurate treatments of gravity, we utilize the effective general relativistic potential in order to streamline calculations, granting us the ability to explore larger sections of parameter space. 
We find that our results utilizing this approximation match very closely the CCSN bounce signal of CFC gravity with GR hydrodynamics \citep{richers:2017}.

The main contributors to the GW signal (10--300 ms postbounce) are postbounce convection, the SASI, and the surface g-modes of the PNS \citep{moro:2018}.  By establishing a positive angular momentum gradient, the convection is suppressed according to the Solberg-H{\o}iland stability criterion \citep{endal:1978,fryer:2000}.  The more oblate shock front inhibits the bipolar sloshing of the SASI.  Since the SASI and convection are the principle drivers exciting the g-modes of the PNS, vibrational emission from the PNS is also inhibited by rotation.  
We, therefore, find that rotation in 2D CCSN simulations results in the muting of GW emission.
This result is consistent across progenitors with different ZAMS masses.

Before the PNS g-mode signal is completely muted, as rotation gradually increases, this signal is pushed to lower peak frequencies and can be characterized by its dynamical frequency.  This observation is no coincidence as both  fundamentally depend on the radius and mass of the PNS.  With more centrifugal support, the PNS has a larger radius.  This larger radius causes the surface of the PNS to emit at lower frequencies, thereby producing a `flatter', lower frequency signal.

We reveal a novel rotational effect on the GW signal during the accretion-phase.  We notice that the nonrotating progenitors all produce low frequency signals ($\sim 40$ Hz) that are below the plausible detection threshold of the aLIGO and KAGRA detectors, whereas the progenitors with larger angular velocities produce measurable GW signals in this frequency range.  We attribute this increase of low frequency emission to the SASI and postshock convection.  For nonrotating progenitors, the convective velocity within the postshock region is high, emitting GWs $\sim 100$ Hz.  As rotational velocity increases, the PNS GW contribution is reduced.  Likewise, as the convection slows, the mass within the gain region emits at lower GW frequencies.  The slower convective flows reduce the total amount of GW power and push the peak GW frequency from the gain region to lower values.  Whereas previous rotating, core collapse, GW studies have focused on the bounce signal as means to determine rotational features, or have focused on late time signals without rotation, our study unifies both facets, and opens the door to measuring GW signals beyond the bounce phase that encode progenitor, angular momentum information. 
We postpone asserting quantitative relations between low frequency emission and progenitor angular momentum until we incorporate more detailed microphysics.

%As with any numerical calculation, resolution has a definite impact on the outcome of a simulation.  To place our simulations in the context of current work in the field, we utilize a 2D cylindrical geometry.  Our maximum outer, radial boundary is $1.0 \times 10^4$ km, with nine levels of refinement, yielding a finest grid spacing of 0.65 km.  The rotating model of \citet{andresen:2018} uses a Yin-Yang grid, with two grid patches of initial resolution  400, 56, and 144 zones in the radial, polar, and azimuthal direction, respectively.  \citet{morozova:2018}  has a finest radial spacing of 0.5 km and a varying radial resolution from $\sim 0.65^o-0.95^o$

While our approximations have allowed us to make large sweeps of parameter space, they leave room for us to include more robust microphysics.  In an ideal situation, we would compute 3D simulations, include full GR, magnetohydrodynamics, and GR Boltzmann neutrino transport that incorporates velocity dependence and inelastic scattering on electrons and nucleons.  These additions would allow for more accurate gravitational waveforms and allow other phenomena to occur, for example the $m\ne 0$ (spiral) modes of the SASI. \citet{andresen:2018} recently highlight the rotational effects on GWs in 3D.  Inherent to the 3D nature of their study, they find the strongest GW amplitudes at high rotation velocities due to these spiral modes.  The 2D geometry of our study, however, allows us to observe the relative strength of the convective signal, without interference from $m\ne 0$ modes, as we extend beyond the case of a single rotational velocity.
While the physical origin of this muting that damps the convection and the SASI is not constrained only to 2D, but in 3D, as \citet{andresen:2018} point out, other non-axisymmetric instabilities can contribute to significant GW emission at late times, negating this rotational muting effect.
Thus, once again, we are reminded of the key role of 3D simulations in the study of the CCSN mechanism.

@@

We would like to thank Jess McIver for pointing us to the aLIGO and AdV sensitivity curves.  M.A.P. was supported under the Michigan State University Distinguished Fellowship. 
S.M.C. is supported by the U.S. Department of Energy, Office of Science, Office of Nuclear Physics,
under Award numbers DE-SC0015904 and DE-SC0017955 and the Chandra
X-ray Observatory under grant number TM7-18005X.
This research was supported by the Exascale Computing Project (17-SC-20-SC), a collaborative effort of two U.S. Department of Energy organizations (Office of Science and the National Nuclear Security Administration) responsible for the planning and preparation of a capable exascale ecosystem, including software, applications, hardware, advanced system engineering, and early testbed platforms, in support of the nation’s exascale computing imperative.
The software used in this
work was in part developed by the DOE NNSA-ASC OASCR Flash Center at
the University of Chicago.

    \software{\\\href{http://flash.uchicago.edu/site/}{FLASH} \citep{fryxell:2000,fryxell:2010},\\ \href{https://matplotlib.org/}{Matplotlib} \citep{hunter:2007},\\
    \href{http://www.nulib.org}{NuLib}
    \citep{oconnor:2015},\\
    \href{http://www.numpy.org/}{NumPy} \citep{vanderwalt:2011},\\ \href{https://www.scipy.org/}{SciPy} \citep{jones:2001}}

>>>>

\author[0000-0002-4983-4589]{Michael A. Pajkos}
\affiliation{Department of Physics and Astronomy, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Department of Computational Mathematics, Science, and Engineering, Michigan State University, East Lansing, MI 48824, USA}
\affiliation{Joint Institute for Nuclear Astrophysics-Center for the Evolution of the Elements, Michigan State University, East Lansing, MI 48824, USA}

@@

\author[0000-0002-1473-9880]{Kuo-Chuan Pan}
\affiliation{Department of Physics and Institute of Astronomy, National Tsing Hua University, Hsinchu 30013, Taiwan}

@@

\section{Methods and Simulation Setup}
\label{sec:method}
We utilize the \texttt{FLASH} (version 4) multiscale, multiphysics adaptive mesh refinement simulation framework for our simulations \citep{fryxell:2000,dubey:2009}.\footnote[7]{\url{http://flash.uchicago.edu/site/}}  We employ a modified, GR, effective potential \citep{marek:2006, oconnor:2018} incorporated into the multipole Poisson solver of \citet{couch:2013a}, where we retain spherical harmonic orders up through 16.   We utilize the SFHo EOS in all of our 15 simulations \citep{steiner:2013}.  Our grid setup is a 2D cylindrical geometry with the PARAMESH (v.4-dev) library for adaptive mesh refinement  \citep{macneice:2000}.  The outer boundary is $10^4$ km in all directions, with nine levels of refinement, yielding a finest grid spacing of about 0.65 km.
The maximum allowed level of refinement is decreased as a function of spherical radius, $r$, in order to maintain a resolution aspect ratio, $\Delta x_i / r$, of about 0.01, corresponding approximately to an ``angular'' resolution of $0.5^{\circ}$ .

@@

\begin{figure*}[t!]
  \centering     %%% not \center
  \includegraphics[width=\textwidth]{tbe6tbe300_combined_M1_long_referee.pdf}
  \caption{ASD plot of all progenitors for all rotation rates from $t_{be}+6$ ms $\rightarrow t_{be}+300$ ms, with an assumed distance of 10 kpc.  The rotational muting of the fundamental PNS g-mode is displayed as the peak frequency ($\sim 800$ Hz) becomes less prevalent, with increasing rotation rate.  Likewise, the low-frequency signals ($\sim40$ Hz) from the gain region become more audible, with increasing rotational velocity.  The damping of the vibrational modes of the PNS allows the slower postshock convection to contribute more to the overall GW signal.  Plotted in the black dashed line is the design sensitivity curve for aLIGO in the zero-detuning, high-sensitivity configuration \citep{barsotti:2018}.  The cyan dashed line is the predicted KAGRA detuned, sensitivity curve \citep{komori:2017}.  The purple dashed line is the design sensitivity curve for AdV \citep{abbott:2018}.}
  \label{fig:spetra_long}
\end{figure*}

@@

Overlaid on our ASD plots is the expected sensitivity of future GW observatories.  
In the black, cyan, and purple dashed lines we have plotted the sensitivity curves of design sensitivity for aLIGO in the zero-detuning, high-sensitivity configuration, the predicted KAGRA detuned sensitivity curve, and design sensitivity for AdV, respectively \citep{komori:2017,abbott:2018,barsotti:2018}.
These curves represent the incoherent sum of the principal noise sources to the best understanding of the respective collaborations.  While these curves do not guarantee the performance of the detectors, they act as good guides for their anticipated sensitivities nonetheless.

Beyond the decreased prevalence of the peak frequency, an interesting trend emerges in Figure \ref{fig:spetra_long} as rotation increases.  
We separate the GW signals by region within the star.  The top row of Figure \ref{fig:spetra_long} corresponds to GWs originating from the inner 50 km of the supernova, and the GW signal in the bottom row originates from radial distances between 50 and 150 km from the supernova center.  In the top row, we note the first peak of emission, around 80 Hz, is independent of rotation.  We point to the bright, higher $v_\mathrm{aniso}$ region in Figure \ref{fig:vaniso} within the first 25 ms postbounce that is present for all rotational velocities.
Focusing on the bottom row, we highlight a noticeable difference in the amplitude of the low-frequency contributions, particularly around 40 Hz.  The nonrotating progenitors have undetectable low-frequency signals for all three detectors, whereas rotating progenitors create measurable signals at low frequencies. 
This enhanced low-frequency signal may provide an observable feature that can help determine progenitor angular momentum information.

The amplitude of low-frequency GWs in the 50--150 km region of the supernova increases with rotational velocity, but this trend does not occur within the inner 50 km.  As such, we restrict the low-frequency GW contribution to the gain region.  We note the two main physical mechanisms in this region correspond to postshock convection and the SASI.  While both mechanisms are reduced in strength due to rotational effects, they do not completely cease.  This fact is displayed in Figure \ref{fig:vaniso}, as the region between 50 and 150 km is nonzero.  For the nonrotating case, the high convective velocities (bright yellow) create higher frequency GWs within the 100 km region of interest.  As rotation velocity increases, convective velocities decrease enough to cease exciting the vibrational modes of the PNS.  These slower convective flows thereby reduce the total amount of power produced by the GWs and push the peak GW frequency---from the gain region---to lower frequencies.  Performing an order-of-magnitude estimate on the source of the low-frequency signal, from Figure \ref{fig:vaniso}, we find  $v_{\mathrm{aniso}} \sim 1 \times 10^9$ cm s$^{-1}$ for $\Omega_0 = 0$ rad s$^{-1}$ and $v_{\mathrm{aniso}} \sim 5 \times 10^8$ cm s$^{-1}$ for $\Omega_0 = 3$ rad s$^{-1}$.  As the region of interest is $\sim 10^7$ cm, we yield an estimated frequency of emission $f_{\mathrm{low}}$ around $\sim 100$ Hz and $\sim 50$ Hz, respectively.  These quantitative frequency estimates are reflected in the ASD as the contribution from peak frequency ($\sim 100$ Hz) from the gain region decreases, while the contribution $\sim 40$ Hz increases.

@@

The strength of this project is its ability analyze GWs hundreds of milliseconds postbounce from multiple progenitors while accurately accounting for rotation and neutrinos.  The wide breadth of parameter space we examine allows us to reveal certain rotational effects on the GW signal in the context of a controlled study.  We have explored the influence of rotation on the GW emission from CCSNe for four different progenitors and four different core rotational speeds.  
We point out that there exists a roughly linear relation between compactness, $\xi$, and the differential rotation parameter, $A$, as defined in Equation (\ref{eq:omega}). 
Using this relation, we calculate appropriate $A$ values for each progenitor mass, based on their individual compactness parameters of the \citet{Suk:2016} progenitors.  Of our 15 simulations, only two nonrotating progenitors have average shock radii that show substantial shock expansion, while the remaining rotating progenitors do not because of rotationally inhibited convection in the gain region and less neutrino production.  In agreement with other recent work \citep[e.g.,][]{summa:2018}, we find a complex interplay between centrifugal support and neutrino heating as successful explosions do not display a monotonic relationship with rotation.

While there are more accurate treatments of gravity, we utilize the effective GR potential in order to streamline calculations, granting us the ability to explore larger sections of parameter space. 
We find that our results utilizing this approximation match very closely the CCSN bounce signal of CFC gravity with GR hydrodynamics \citep{richers:2017}.

The main contributors to the GW signal (10--300 ms postbounce) are postbounce convection, the SASI, and the surface g-modes of the PNS \citep{moro:2018}.  By establishing a positive angular momentum gradient, the convection is suppressed according to the Solberg-H{\o}iland stability criterion \citep{endal:1978,fryer:2000}.  The more oblate shock front inhibits the bipolar sloshing of the SASI.  Since the SASI and convection are the principal drivers exciting the g-modes of the PNS, vibrational emission from the PNS is also inhibited by rotation.  
We, therefore, find that rotation in 2D CCSN simulations results in the muting of GW emission.
This result is consistent across progenitors with different ZAMS masses.

Before the PNS g-mode signal is completely muted, as rotation gradually increases, this signal is pushed to lower peak frequencies and can be characterized by its dynamical frequency.  This observation is no coincidence as both  fundamentally depend on the radius and mass of the PNS.  With more centrifugal support, the PNS has a larger radius.  This larger radius causes the surface of the PNS to emit at lower frequencies, thereby producing a ``flatter,'' lower frequency signal.

We reveal a novel rotational effect on the GW signal during the accretion-phase.  We notice that the nonrotating progenitors all produce low-frequency signals ($\sim 40$ Hz) that are below the plausible detection threshold of the aLIGO and KAGRA detectors, whereas the progenitors with larger angular velocities produce measurable GW signals in this frequency range.  We attribute this increase of low-frequency emission to the SASI and postshock convection.  For nonrotating progenitors, the convective velocity within the postshock region is high, emitting GWs $\sim 100$ Hz.  As rotational velocity increases, the PNS GW contribution is reduced.  Likewise, as the convection slows, the mass within the gain region emits at lower GW frequencies.  The slower convective flows reduce the total amount of GW power and push the peak GW frequency from the gain region to lower values.  Whereas previous rotating core-collapse GW studies have focused on the bounce signal as a means to determine rotational features, or have focused on late time signals without rotation, our study unifies both facets and opens the door to measuring GW signals beyond the bounce phase that encode progenitor, angular momentum information. 
We postpone asserting quantitative relations between low-frequency emission and progenitor angular momentum until we incorporate more detailed microphysics.

While our approximations have allowed us to make large sweeps of parameter space, they leave room for us to include more robust microphysics.  In an ideal situation, we would compute 3D simulations, including full GR, magnetohydrodynamics, and GR Boltzmann neutrino transport that incorporates velocity dependence and inelastic scattering on electrons and nucleons.  These additions would allow for more accurate gravitational waveforms and allow other phenomena to occur, for example the $m\ne 0$ (spiral) modes of the SASI. \citet{andresen:2018} recently highlighted the rotational effects on GWs in 3D.  Inherent to its 3D nature, their study finds the strongest GW amplitudes at high rotation velocities due to these spiral modes.  The 2D geometry of our study, however, allows us to observe the relative strength of the convective signal, without interference from $m\ne 0$ modes, as we extend beyond the case of a single rotational velocity.
While the physical origin of this muting that damps the convection and the SASI is not constrained only to 2D, in 3D, as \citet{andresen:2018} point out, other nonaxisymmetric instabilities can contribute to significant GW emission at late times, negating this rotational muting effect.
Thus, once again, we are reminded of the key role of 3D simulations in the study of the CCSN mechanism.

@@

We would like to thank Jess McIver for pointing us to the aLIGO and AdV sensitivity curves.  M.A.P. was supported by a Michigan State University Distinguished Fellowship. 
S.M.C. is supported by the U.S. Department of Energy, Office of Science, Office of Nuclear Physics,
under award Nos. DE-SC0015904 and DE-SC0017955 and the \textit{Chandra
X-ray Observatory} under grant No. TM7-18005X.
This research was supported by the Exascale Computing Project (17-SC-20-SC), a collaborative effort of two U.S. Department of Energy organizations (Office of Science and the National Nuclear Security Administration) that are responsible for the planning and preparation of a capable exascale ecosystem, including software, applications, hardware, advanced system engineering, and early testbed platforms, in support of the nation's exascale computing imperative.
The software used in this
work was in part developed by the DOE NNSA-ASC OASCR Flash Center at
the University of Chicago.

    \software{FLASH (see footnote 7) \citep{fryxell:2000,fryxell:2010}, Matplotlib\footnote[8]{\url{https://matplotlib.org/}} \citep{hunter:2007},
    NuLib\footnote[9]{\url{http://www.nulib.org}}
    \citep{oconnor:2015},
    NumPy\footnote[10]{\url{http://www.numpy.org/}} \citep{vanderwalt:2011}, SciPy\footnote[11]{\url{https://www.scipy.org/}} \citep{jones:2001}}
